---
title: "R Markdown Training"
author: "Sally Thompson"
date: "28 July 2022"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: yes
    
---

```{r setup, include = FALSE}

# include = FALSE: ensures that code and results don't appear in the knitted document. 
# So when you knit the document this chunk will not appear.

# This is a global option, so this will apply to all of the code chunks in the document

knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE, 
                      echo = FALSE)
# change the echo to TRUE to show the R code in the document

n <- 0
```

# Introduction

This guide has been written for you to follow along, you will create your own 
version of this file as you go. It will work best if you have two screens - have 
this document open on one screen, and R Studio with the code and output on the 
other (main) screen.  

Before we go any further, copy the code below into your console and run it. This 
will identify and install any missing packages that are required for this script to run correctly.  

``` {r, eval=FALSE, echo = TRUE}

## If a package is installed, it will be loaded. If any 
## are not, the missing package(s) will be installed from CRAN

## First specify the packages of interest
packages = c("dplyr", "tidyr", "lubridate", "zoo",
             "ggplot2", "DT", "reactable", "plotly")

## Now installs if required
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      }
  }
)

```

Start by opening the Rmd file "developing script.Rmd" in R Studio. This is an unformatted 
version of the markdown file that created this guide. You run a markdown by 'knit'ing 
it (look for the icon - it's a ball of wool with a knitting needle in it). Click 
the 'knit' button and look at the document it creates. How does it compare to this 
file? (The first time you knit an Rmd script you may receive warning messages about 
packages that are required but aren't yet installed. We have tried to pre-empt 
this by running the code above, but if it still happens install those packages 
and knit again.)  

As you work through this guide you will add formatting and content to your script 
until your knitted document looks like this one! 

This guide does not go into the magic that happens between hitting the 'knit' 
button and the output file being produced. There are plenty of resources on the 
internet if you want to explore this further. For more details about all things markdown visit <http://rmarkdown.rstudio.com>.  

This guide will start by looking at formatting your document, then will add in 
some data and charts, and improve how they look. The final section adds some 
interactivity to some of the charts.


# What is R Markdown?

Markdown is a way to create and format documents, the types of which include 
(but aren't limited to) html, pdf, Word, dashboards, slides etc. Two major benefits 
of creating these using R Markdown are reproducibility and automation. Code can 
be embedded, so the creation of properly formatted text plus charts and visualisations 
all happen with one click. This training guide will focus exclusively on producing 
html output, but hopefully by the end of it you will have the confidence and skills 
to investigate other output formats.   

# Formatting

## You will no doubt be familiar with headers.

### They can get smaller
#### and smaller
##### and even smaller still

R markdown uses the # symbol to create headers. A level one header uses one #, 
use ## for a level two header etc. Note how, in the script,  the background colour 
of the row changes.

**Exercise `r n<-n+1; n`:** Turn 'Introduction', 'What is R Markdown?' and 'Formatting' 
into level 1 headers. Apply subsequent header levels to the text that follows 'Formatting'.  

## Italics, bold and super/sub scripts

The asterisk is used to make text bold or italic (or both). For italics, apply one * 
either side of the word/phrase, for bold use ** either side of the word/phrase. 
If you want both italics **and** bold, apply ***.  
In the same way, apply ^ for superscript and ~~ for strike-through (note the 2 ~~ symbols).  

**Exercise `r n<-n+1; n`:** 

*Make this text italic *   

**make this text bold**  

***Really emphasise this by making it bold and italics***  

Apply a superscript to ^part^ of this text  

~~Strike through this text~~  

## Lists

Just as in Word, you can create bulleted and numbered lists. For bullet points, 
use *, - or + at the start of each point to be bulleted. Indent and use + for sub-bullets.  

**Exercise `r n<-n+1; n`a:** turn the first list into a bulleted list with sub-bullet, 
and the second list into a numbered list.

* Asterisk, dash or + sign for bullet points
- and other lists
    + indent for sub-lists
+ and more points if that's what you want  

**Exercise `r n`b:** Create numbered lists by starting each line with (@). The 
numbers will dynamically adjust.

(@) markdown does the numbering for you
(@) so there's no chance of mis-counting after an edit
(@) it will automatically update
(@) isn't that clever!  

You **can** use numbers to make a numbered list, but you need to add a full-stop 
after each number.  

1. you can choose to number lists yourself
2. it must start at number 1,
5. but after that it will number consecutively even if you wanted this item to 
be numbered e.g. 5  


## Paragraphs & Spacing

To create a line break you have to double space after the last sentence then press 
return.

Even if you press enter in the markdown script, it won't appear as a new paragraph 
in the document unless you end the sentence with a double space.
This line looks like it should be a new paragraph in the markdown code, but it 
isn't once it renders to html.  

> Make text stand out by using block quotes  
> Preceed each line with >   
> This can be useful for headline statements  
> But remember to put two spaces after each sentence 

**Exercise `r n<-n+1; n`:** turn the previous four lines into block code  

Use 3 asterisks to create a line break below this sentence.

***
The asterisk is turning out to be quite a versatile symbol!  

## Links

A useful feature for any document that will be viewed electronically is linking 
to websites or email addresses. To display the complete URL, wrap it in angled 
brackets <>. A good starting point for learning more about R Markdown is <http://rmarkdown.rstudio.com>.  

Instead, you might want a clickable link that displays some alternative text. 
This uses a combination of square and round brackets - wrap the readable text in 
square brackets, and follow it with the URL link in round brackets. [R Markdown: the definitive guide](https://bookdown.org/yihui/rmarkdown/) is another really useful resource. 
This is the same method used for creating a clickable email link: ` 
[text to display](mailto:email@email.address)`

**Exercise `r n<-n+1; n`:**   

(@) Contact: (add some alt text and your email address as a hyperlink)  

(@) Go back to the Introduction section and make the link to rmarkdown active.  

(@) We are going to be using data from the NHS Scotland Open Data platform. Create 
some alt text and link it to opendata.nhs.scot 

(@) Knit and check your links work.

## Appearance

Your document should be looking more like this one, but it's still missing the 
table of contents. We add that in the YAML header. While we are there, we will 
change the theme. Choose from any of the [Bootswatch](https://bootswatch.com/3/) 
free themes, or download other templates.  

**Exercise `r n<-n+1; n`:** amend your YAML header as below. Note the indentation 
after 'output', and further indentation after 'html_document' - in YAML, this is 
essential so it knows to link it all back to the html document output. See if you 
like any of the other free themes from [Bootswatch](https://bootswatch.com/3/) 
(cerulean, cosmo, cyborg, darkly, flatly, journal, lumen, paper, readable, sandstone, 
simplex, slate, spacelab, superhero, united or yeti).  


```
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: yes
```

While you are in the YAML header, change the author and date too.  


# Dealing with data

Now things start to get really interesting! We are going to import some data, 
do some wrangling and create some outputs to display. We will be using R to do 
all this, but R Markdown supports other languages too.  

There are two ways of inserting code into a document: blocks of code are wrapped 
up in a code chunk, but you can also insert objects inline in text. You create a 
code chunk using the `Insert` option at the top of the pane, the keyboard shortcut
`Ctrl + Alt + I` or manually, by wrapping it in three backticks at the start and 
end of the chunk. You also need to use curly brackets to declare which language the code is in.


````markdown
`r ''````{r add-chunk-name}

# insert R code here

`r ''````

````

It's good practice to name your code chunks (each chunk needs a unique name), 
keep it short but descriptive of what it is doing. Don't use spaces, dots or 
underscores in chunks - use "-" if you need to use a separator.

Adding chunk options within the curly brackets can e.g. hide the code, only show 
the results, evaluate the code or not (plus many more options). If you want the 
same rules to apply to all code chunks you would include it in the global options. 
Look at the very first code chunk in the script - `include = FALSE` excludes the 
code chunks from the output by default.  

Use single backticks to run code inline with text. You still need to include `r` 
but this time it doesn't need to be in curly brackets. For example, to print today's 
date use `` `r
format(Sys.time(), '%d %B %Y')` `` inline with the rest of your text.  

**Exercise `r n<-n+1; n`:**  

(@) If you don't already have the package installed, run 
`remotes::install_github("Public-Health-Scotland/phsopendata", upgrade = "never")` 
in the console. This will install the PHS Open Data package, which we will use to 
import some data.  

(@) Create a code chunk, name it and paste the following code into the chunk.


```{r import-data, echo = TRUE}

## libraries
library(dplyr)
library(tidyr)
library(phsopendata)    # easily access open data from PHS
library(lubridate)      # parse dates correctly
library(zoo)

## one way to get data from the PHS opendata repository (https://www.opendata.nhs.scot/) is by the resource id, found in the metadata section

# importing the last 12 months of monthly A&E data
monthly_ae <- get_resource(res_id =	"2a4adc0a-e8e3-4605-9ade-61e13a85b3b9") %>% 
  mutate(month_end = ym(Month)) %>% 
  filter(between(month_end, max(month_end)- months(12), max(month_end))) %>% 
  select(TreatmentLocation, DepartmentType, month_end, NumberOfAttendancesAggregate, NumberMeetingTargetAggregate)

# grouping for use later on
monthly_ae_gp <- monthly_ae %>%
  group_by(DepartmentType, month_end) %>% 
  summarise("Number Of Attendances" = sum(NumberOfAttendancesAggregate), 
            "Number Meeting Target" = sum(NumberMeetingTargetAggregate)) %>% 
  pivot_longer(
    cols = starts_with("Number"),
    names_to = "type",
    values_to = "count"
  ) %>% 
  ungroup()

```


When you knit, it doesn't look like anything has happened. But behind the scenes, the data
has been imported and filtered, and the object `monthly_ae` is waiting to be used.  

In the next code chunk we will create some simple plots, and also introduce tabsets. 
These are a
useful way to include lots of content without the document becoming excessively long.  


## Monthly Attendances at A&E {.tabset .tabset-fade .tabset-pills} 

To create tabsets add {.tabset} to the header row. Then create two (or more, how 
many tabs do you want?) headers at the next level down, with a code chunk in each section.


### Attendances at EDs

```{r plot-ed}

library(ggplot2)

pl_colours <- c("darkred", "steelblue")

plot1 <-ggplot(monthly_ae_gp %>% filter(DepartmentType == "Emergency Department"), 
               aes(x = month_end, y = count, group = type, colour = type)) +
  geom_line() +
 expand_limits(y = 0) +
 labs(title = "Number of Attendances and 4 hour target at Emergency Departments",
      x = "month ending",
      y = "number of Attendances") +
  scale_colour_manual(values = pl_colours) +
 theme_minimal()

 plot1

```


You can add commentary before the next section heading.


### Attendances at MIUs

``` {r plot-miu}

plot2 <-ggplot(monthly_ae_gp %>% filter(DepartmentType == "Minor Injury Unit or Other"), 
               aes(x = month_end, y = count, group = type, colour = type)) +
  geom_line() +
 expand_limits(y = 0) +
 labs(title = "Number of Attendances and 4 hour target at Minor Injury Units",
      x = "month ending",
      y = "number of Attendances") +
  scale_colour_manual(values = pl_colours) +
 theme_minimal()

plot2

```

And different commentary for this chart.

## {.unlisted .unnumbered}

You aren't limited to just displaying charts in tabsets - you might want to also 
include the data behind a chart, in a table. 

**Exercise `r n<-n+1; n`:**

1. add `{.tabset}` to the *Monthly Attendances at A&E* level 2 header.  
2. create a level 3 header, *Attendances at EDs*  
3. paste the following code into a code chunk. Add some suitable commentary 
below the code chunk  

``` {r plot-ed-code, echo = TRUE, eval = FALSE}

library(ggplot2)

pl_colours <- c("darkred", "steelblue")

plot1 <-ggplot(monthly_ae_gp %>% filter(DepartmentType == "Emergency Department"), 
               aes(x = month_end, y = count, group = type, colour = type)) +
  geom_line() +
 expand_limits(y = 0) +
 labs(title = "Number of Attendances and 4 hour target at Emergency Departments",
      x = "month ending",
      y = "number of Attendances") +
  scale_colour_manual(values = pl_colours) +
 theme_minimal()

 plot1
 
```  
 
4. repeat with another level 3 header, titled *Attendances at MIUs*  


```{r plot-miu-code, echo = TRUE, eval = FALSE}

plot2 <-ggplot(monthly_ae_gp %>% filter(DepartmentType == "Minor Injury Unit or Other"), 
               aes(x = month_end, y = count, group = type, colour = type)) +
  geom_line() +
 expand_limits(y = 0) +
 labs(title = "Number of Attendances and 4 hour target at Minor Injury Units",
      x = "month ending",
      y = "number of Attendances") +
  scale_colour_manual(values = pl_colours) +
 theme_minimal()

plot2

```

5. If you want to make it look a bit snazzy, add a bit of flair to the tabsets 
by adding `.tabset-fade` and/or `.tabset-pills` to `{.tabset}` (keep it all inside the curly brackets).

# Housekeeping

By now the script is getting quite long, and can be tricky to navigate. If you've 
been good at defining sections then you can toggle to show the document outline 
on the right-hand side of the script pane. Other points of good practice:

- load all the libraries you need in the first code chunk. (I haven't done that 
in this document, as I wanted you to see which package is used in each example)  

- similarly, load and wrangle the data early on in the script too, before you start 
creating content. Then the code chunks within the main text can be kept sparse, 
such as just calling a plot.  

- run each chunk in RStudio as you create it to check it works - it's easier to 
troubleshoot this way, than following an error message created when trying to knit.  

- if there's a lot of processing to be done then create this in another R script 
instead, and 'call' it into the markdown by using `source`.  

- be very careful using filepaths - the 'start point' of filepaths can be different 
when running a chunk locally than when knitting. The easiest way to get round that 
is to use the package `here`, which always starts at the location of the project directory.  

# Adding some interactivity

It's all well and good being able to print a chart or table, but it's still a static 
object. Wouldn't it be great if readers could interact with those objects? The 
beauty of rendering the markdown script to html is that we can! What follows are 
just a few examples of packages that can add some interactivity to your reports. 
There aren't any exercises, but follow along with the script to see how to apply 
them. In your script the following code chunks are currently set to not run when 
you knit - delete the code `eval = FALSE` from each code chunk (or change it to 
`TRUE`) so that the code does run from now on.   

## DT (Datatable)

This default output for displaying tables using DT is shown below, but you can 
customise it (add options for exporting the data, change how many rows are displayed, 
plus lots more). However, you can see it doesn't like the `yearmon` date format, 
and displays it as a numeric. (There is probably a workaround...)

```{r datatable}

library(DT)

datatable(monthly_ae %>% 
  rename("Number Of Attendances" = NumberOfAttendancesAggregate,
         "Number Meeting Target" = NumberMeetingTargetAggregate) %>% 
  mutate(month_end = as.yearmon(month_end)))

```

## Reactable

With reactable, you can group data without having to hard code it first, using 
the built-in aggregate functions or define your own. Clicking on a grouped item will unroll it.   

```{r reactable}

library(reactable)

reactable(monthly_ae %>% 
            rename(Location = TreatmentLocation,
                   "Number Of Attendances" = NumberOfAttendancesAggregate,
                   "Number Meeting Target" = NumberMeetingTargetAggregate),
          groupBy = c("DepartmentType", "Location"),
          columns = list(
            month_end = colDef(cell = function(value) strftime(value, "%Y-%b")),
            "Number Of Attendances" = colDef(aggregate = "sum"),
            "Number Meeting Target" = colDef(aggregate = "sum")
          )
)


```

## Plotly

We can apply all sorts of tricks to make our ggplots look fancy, but at the end 
of the day they are still just static plots. Plotly can bring them to life.  Hovering
over the lines will display data points, and you can zoom into a specific area, 
download the plot, plus other options available through the buttons top-right of the plot.. 

```{r plotly}

library(plotly)

plot3 <-monthly_ae_gp %>% filter(type == "Number Of Attendances") %>% 
  ggplot(., aes(x = month_end, y = count, group = DepartmentType, colour = DepartmentType)) +
  geom_line() +
 expand_limits(y = 0) +
 labs(title = "Number of Attendances at A&E",
      x = "month ending",
      y = "number of Attendances") +
  scale_colour_manual(values = pl_colours) +
 theme_minimal()

ggplotly(plot3)

```

Plotly may not be able to render more advanced ggplots using the `ggplotly` function, 
but you can build plots from scratch in plotly. The language and terminology is 
different than ggplot, but ultimately it gives you a lot more functionality.



# What Next?

This is only a brief introduction to R Markdown, there is so much more to explore! 
The suggestions below are outwith the scope of this guide, and are only a tiny 
fraction of what is possible.  

* Automate the same report for multiple groups (e.g. ICSs, LAs, hospitals, ...). 
These are done by defining the parameters. Chapter 15 in [R Markdown: the definitive guide](https://bookdown.org/yihui/rmarkdown/parameterized-reports.html) describes 
this in more detail. This method can also be applied to other types of outputs, 
such as pdf, Word or Powerpoint documents.  

* [Flexdashboard](https://rmarkdown.rstudio.com/flexdashboard/) is an extension 
package to markdown that allows you to build dashboards in html. These are useful 
to display data visualisations, and include other components such as data tables, 
interactive maps, gauges, value boxes etc. It can be made up of multiple pages, 
or you can control how analysis and its conclusions are unveiled through the use of a storyboard.  

* Further customise the appearance of the html document by adding CSS styles and 
html formatting. For instance (look at the source code to see how this effect is created):  

<style>
div.phsblu50 { background-color:#80bcea; border-radius: 5px; padding: 20px;}
</style>
<div class = "phsblu50">

you can change the background colour of a block to further highlight key statements

- and change the <span style="color: #c73918;">**font colour**</span> for specific words/sentences

- making your conclusions really stand out

</div>
